Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-08-18 prop_letter.tex
prop_letter.tex \newcommand{\nb}{\nolinebreak[1]\ }
expectation values

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{framed}
\newcommand{\amr}[1]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{green}{Amr says: #1}\end{minipage}\end{framed}}
\newcommand{\yutsung}[1]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{purple}{Yu-Tsung says: #1}\end{minipage}\end{framed}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}~\middle|~{#2}\right\} }}
\newcommand{\Tr}{\ensuremath{\mathop{\mathrm{Tr}}\nolimits}}
\allowdisplaybreaks
\newcommand{\coreBorn}{\ensuremath{\overline{\Hilb}}}
\newcommand{\mul}[1][]{\ensuremath{\mu^{\mathrm{L{#1}}}}}
\newcommand{\mur}[1][]{\ensuremath{\mu^{\mathrm{R{#1}}}}}
\setcounter{secnumdepth}{3}
% https://tex.stackexchange.com/questions/345284/can-i-have-an-almost-non-breaking-space-in-latex
% Change non-breaking space to \nolinebreak[1] 
% Hence, it will not hyphenate too much surrounding words.
\newcommand{\nb}{\nolinebreak[1] }
\newcommand{\rmd}{\mathrm{d}}

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$,
the identity projector~$\mathbb{1}$, projectors of the form
$\proj{\phi}$ where $\ket{\phi}$ is a pure quantum state (an element
of a Hilbert space), sums of \emph{orthogonal} projectors $P_0$
and~$P_1$ with $P_0P_1=\mathbb{0}$, and products of \emph{commuting}
projectors $P_0$ and~$P_1$ with $P_0P_1=P_1P_0$. In a well-defined
quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$
is the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given an observable~$\mathbf{O}$ with spectral decomposition
$\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$ one can define the 
expectation value of that observable as
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=1}^{N}\lambda_{i}\mu(P_{i})\,.\label{eq:quantum-expectation}
\end{equation}

A conventional quantum probability measure can be easily
constructed if one knows the current state of the quantum system by
using the Born rule.  Specifically, for each pure normalized quantum
state $\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$. Two fragments of
valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $\ket{\phi_1}=(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the second
corresponds to the quantum system being in the pure state
$\ket{\phi_2}=(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. The expectation value of the observable $\mathbf{O}$, 
$\expval{\mathbf{O}}_{\mu_{1,2}}$, in
the first case is 1.5 and in the second it is 2.

@@

In the remainder of this section, we will show that our definition of
QIVPMs is consistent with classical IVPMs in the sense that a
restriction of QIVPMs to commuting events recovers the definition of
classical IVPMs. To formalize this connection between QIVPMs and
classical IVPMs, we first note that in contrast to the case with
conventional quantum probability spaces, there is generally no unique
state~$\rho$ consistent with a QIVPM. We say that a QIVPM $\bar{\mu}$
is consistent with a state $\rho$ and projector $P$ if the interval
$\bar{\mu}(P)$ contains the exact probability calculated by the Born
rule, i.e, if:
\begin{equation}
\mu_{\rho}^{\mathrm{B}}\left(P\right)\in\bar{\mu}\left(P\right)
\end{equation}
where $\mu_{\rho}^{\mathrm{B}}\left(P\right)=\Tr\left(\rho P\right)$
is the probability that the event~$P$ is determined by the Born
rule~\citep{Born1983,peres1995quantum,544199}.

@@

\begin{thm}\label{thm:Shapley}
  For a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, if a subspace
  of events~$\events_C\subseteq\events$ commutes, then
  $\coreBorn\left(\bar{\mu},\events_C\right)\ne\emptyset$.
\end{thm}

@@

Since Thm.~\ref{thm:Shapley} corresponds to classical Shapley's
theorem, Thm.~\ref{thm:Rosenmuller} has its quantum counterpart.

\begin{thm} Given a Hilbert space $\Hilb$ of dimension~$d\ge3$,
for every QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, and an
observable~$\mathbf{O}$ with its spectral decomposition~(\ref{eq:spectrum-decomposition}),
let $\events'$ be the minimal subspace of events containing $\left\{ P_{i}^{-}\right\} _{i=1}^{N^{-}}\cup\left\{ P_{i}^{+}\right\} _{i=1}^{N^{+}}$,
then 
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho}\right]\,.\label{eq:quantum-interval-expectation-core}
\end{equation}
\end{thm}

@@

>>>>

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{framed}
\newcommand{\amr}[1]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{green}{Amr says: #1}\end{minipage}\end{framed}}
\newcommand{\yutsung}[1]{\begin{framed}\begin{minipage}{0.9\linewidth}\color{purple}{Yu-Tsung says: #1}\end{minipage}\end{framed}}
\newcommand{\set}[2]{\ensuremath{\left\{ {#1}~\middle|~{#2}\right\} }}
\newcommand{\Tr}{\ensuremath{\mathop{\mathrm{Tr}}\nolimits}}
\allowdisplaybreaks
\newcommand{\coreBorn}{\ensuremath{\overline{\Hilb}}}
\newcommand{\mul}[1][]{\ensuremath{\mu^{\mathrm{L{#1}}}}}
\newcommand{\mur}[1][]{\ensuremath{\mu^{\mathrm{R{#1}}}}}
\setcounter{secnumdepth}{3}
% https://tex.stackexchange.com/questions/345284/can-i-have-an-almost-non-breaking-space-in-latex
% Change non-breaking space to \nolinebreak[1] 
% Hence, it will not hyphenate too much surrounding words.
\newcommand{\nb}{\nolinebreak[1]\ }
\newcommand{\rmd}{\mathrm{d}}

@@

A \emph{probability space} is a mathematical abstraction specifying
the necessary conditions for reasoning coherently about collections of
uncertain
events~\cite{Kolmogorov1950,544199,Griffiths2003,Grabisch2016}. In the
quantum case, the events of interest are specified by \emph{projection
  operators} $P$. These include the empty projector~$\mathbb{0}$,
the identity projector~$\mathbb{1}$, projectors of the form
$\proj{\phi}$ where $\ket{\phi}$ is a pure quantum state (an element
of a Hilbert space), sums of \emph{orthogonal} projectors $P_0$
and~$P_1$ with $P_0P_1=\mathbb{0}$, and products of \emph{commuting}
projectors $P_0$ and~$P_1$ with $P_0P_1=P_1P_0$. In a well-defined
quantum probability
space~\cite{10.2307/2308516,gleason1957,Redhead1987-REDINA,Maassen2010},
each event~$P_{i}$ is mapped to a probability~$\mu(P_{i})$ using a
probability measure~$\mu:\events\rightarrow[0,1]$, where $\events$
is the set of all projectors on a given Hilbert space, subject to the
following constraints: $\mu(\mathbb{0})=0$, $\mu(\mathbb{1})=1$,
$\mu\left(\mathbb{1}-P\right)=1-\mu\left(P\right)$, and for each pair
of \emph{orthogonal} projectors $P_{0}$ and $P_{1}$:
\begin{equation}
{\mu}\left(P_{0}+P_{1}\right)={\mu}\left(P_{0}\right)+{\mu}\left(P_{1}\right)\,.\label{eq:QuantumProbability-Addition}
\end{equation}
Given an observable~$\mathbf{O}$ with spectral decomposition
$\mathbf{O}=\sum_{i=1}^{N}\lambda_{i}P_{i}$ one can define the 
expectation value of that observable as:
\begin{equation}
\expval{\mathbf{O}}_{\mu}=\sum_{i=1}^{N}\lambda_{i}\mu(P_{i})\,.\label{eq:quantum-expectation}
\end{equation}

A conventional quantum probability measure can be easily
constructed if one knows the current state of the quantum system by
using the Born rule.  Specifically, for each pure normalized quantum
state~$\ket{\phi}$, the Born rule induces a probability measure
$\mu_{\phi}^{\mathrm{B}}$ defined as
$\mu_{\phi}^{\mathrm{B}}(P)=\ip{\phi}{P\phi}$. The situation
generalizes to mixed states
$\rho = \sum_{j=1}^{N}q_{j}\proj{\phi_{j}}$, where
$\sum_{j=1}^{N}q_{j}=1$ in which case the generalized Born rule
induces a probability measure $\mu_{\rho}^{\mathrm{B}}$ defined as
$\mu_{\rho}^{\mathrm{B}}\left(P\right) = \Tr\left(\rho P\right) =
\sum_{j=1}^{N}
q_{j}\mu_{\phi_{j}}^{\mathrm{B}}\left(P\right)$~\cite{peres1995quantum,544199,Jaeger2007}.

As an example, consider a three-dimensional Hilbert space with
orthonormal basis $\{\ket{0},\ket{1},\ket{2}\}$ and an observable
$\mathbf{O}$ with spectral decomposition
$\mathbf{O} = \proj{0} + 2\,\proj{1} + 3\,\proj{2}$. Two fragments of
valid probability measures $\mu_1$ and $\mu_2$ that can be associated
with this space are: $\mu_1(\proj{0}) = 0.5$, $\mu_1(\proj{1}) = 0.5$,
$\mu_1(\proj{2}) = 0$, and $\mu_2(\proj{0}) = 0.5$,
$\mu_2(\proj{1}) = 0$, $\mu_2(\proj{2}) = 0.5$. By the Born rule, the
first probability measure corresponds to the quantum system being in
the pure state $\ket{\phi_1}=(\ket{0}+\alpha\ket{1})/\sqrt{2}$ and the
second corresponds to the quantum system being in the pure state
$\ket{\phi_2}=(\ket{0}+\alpha\ket{2})/\sqrt{2}$, where in both cases
$|\alpha|=1$. The expectation values of the observable $\mathbf{O}$,
$\expval{\mathbf{O}}_{\mu_{1,2}}$, are 1.5 in the first case is 1.5
and 2 in the second.

@@

In the remainder of this section, we will show that our definition of
QIVPMs is consistent with classical IVPMs in the sense that a
restriction of QIVPMs to commuting events recovers the definition of
classical IVPMs. To formalize this connection between QIVPMs and
classical IVPMs, we first note that in contrast to the case with
conventional quantum probability spaces, there is generally no unique
state~$\rho$ consistent with a QIVPM. We say that a QIVPM $\bar{\mu}$
is consistent with a state $\rho$ and projector~$P$ if the interval
$\bar{\mu}(P)$ contains the exact probability calculated by the Born
rule, i.e, if:
\begin{equation}
\mu_{\rho}^{\mathrm{B}}\left(P\right)\in\bar{\mu}\left(P\right)
\end{equation}
where $\mu_{\rho}^{\mathrm{B}}\left(P\right)=\Tr\left(\rho P\right)$
is the probability that the event~$P$ is determined by the Born
rule~\citep{Born1983,peres1995quantum,544199}.

@@

\begin{thm}\label{thm:Shapley}
  For QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, if a subspace
  of events~$\events_C\subseteq\events$ commutes, then
  $\coreBorn\left(\bar{\mu},\events_C\right)\ne\emptyset$.
\end{thm}

@@

In conventional quantum mechanics the expectation value of an
observable is defined in Eq.~\ref{eq:quantum-expectation} as a precise
real number. The generalization of quantum measures to QIVPMs implies
that the real-valued expectation value will be generalized to an
interval.

\begin{definition}[Expectation Value of Observables over QIVPMs] Let
  $\mathscr{I}$ be a set of intervals; $\Hilb$ a Hilbert space of
  dimension~$d\ge3$ with event space $\events$; and $\mathbf{O}$ an
  observable with spectral decomposition
  $\Sigma_{i=1}^N \lambda_iP_i$. Let $\events'$ be the minimal
  subspace of events containing all the projectors $P_i$, then we define:
\begin{equation}
\expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}
\expval{\mathbf{O}}_{\mu_{\rho}^{\mathrm{B}}}\right]\,.\label{eq:quantum-interval-expectation-core}
\end{equation}
\end{definition}

If $\bar{\mu}$ is a conventional (Born) probability measure defined by
state $\rho$, then $\coreBorn\left(\bar{\mu},\events'\right)$ will
contain the singleton state $\rho$ and both endpoints of the interval
will agree, thus reducing the definition to that of
Eq.~\ref{eq:quantum-expectation}. We also note that, when restricted
to commuting projectors, the definition above is consistent with the
classical notion of the \emph{Choquet
  integral}~\cite{GilboaSchmeidler1994,Grabisch2016} which is used to
calculate the expectation value of random variables as a weighted
average.

@@

% \begin{thm} Given a Hilbert space $\Hilb$ of dimension~$d\ge3$,
% for every QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$, and an
% observable~$\mathbf{O}$ with its spectral decomposition~(\ref{eq:spectrum-decomposition}),
% let $\events'$ be the minimal subspace of events containing $\left\{ P_{i}^{-}\right\} _{i=1}^{N^{-}}\cup\left\{ P_{i}^{+}\right\} _{i=1}^{N^{+}}$,
% then 
% \begin{equation}
% \expval{\mathbf{O}}_{\bar{\mu}}=\left[\min_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho},\max_{\rho\in\coreBorn\left(\bar{\mu},\events'\right)}\expval{\mathbf{O}}_{\rho}\right]\,.\label{eq:quantum-interval-expectation-core}
% \end{equation}
% \end{thm}
